---
title: "GLS"
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 2
date: "2025-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readxl)
data <- read_excel("dataset.xlsx") 
```

```{r}
colnames(data)
```

# linear regression

```{r}

model <- lm(sexual_crime ~ County+poor_mental_health_days+excessive_drinking_rate+adult_smoking_rate+unemployment_rate+median_household_income+uninsured+mental_health_providers, data = data)
summary(model)
```

4 significant: Adult_smoking_rate + unemployment_rate + uninsured + mental.health.providers

**diagnostics**

```{r}
library(lmtest)
dwtest(model) # true autocorrelation is greater than 0
```

```{r}
data$Time <- as.numeric(as.factor(data$Time))
plot(x = data$Time,y = model$residuals)
```

```{r}
plot(model) # increasing variance
```



# 1a: GLS AR1

```{r}
library(tidyverse)
data$Time <- as.numeric(as.factor(data$Time))
library(nlme)
mod_gls_1a <- gls(sexual_crime ~ County + poor_mental_health_days + excessive_drinking_rate +
               adult_smoking_rate + unemployment_rate + 
               uninsured + mental_health_providers, 
               data = data, 
               correlation = corAR1(form = ~ Time | County), 
               method = "ML")
summary(mod_gls_1a)
```

0 variable significant

**diagnostics**

```{r}
par(mfrow = c(1, 2))
acf(mod_gls_1a$residuals, main = "ACF of GLS Residuals")
pacf(mod_gls_1a$residuals, main = "PACF of GLS Residuals")
```

```{r}
plot(mod_gls_1a$residuals ~ data$Time,
     xlab = "Time", ylab = "residuals")
```

```{r}
plot(mod_gls_1a) # increasing variance
```



# 1b ARMA(1,1)

```{r}
library(nlme)
mod_gls_1b <- gls(sexual_crime ~ County + poor_mental_health_days + excessive_drinking_rate +
               adult_smoking_rate + unemployment_rate + 
               uninsured + mental_health_providers, 
               data = data, 
               correlation = corARMA(p = 1, q = 1, form = ~ Time | County), 
               method = "ML")
summary(mod_gls_1b)
```

0 variable significant

**diagnostics**

```{r}
par(mfrow = c(1, 2))
acf(mod_gls_1b$residuals, main = "ACF of GLS Residuals")
pacf(mod_gls_1b$residuals, main = "PACF of GLS Residuals")
```

```{r}
plot(mod_gls_1b$residuals ~ data$Time,
     xlab = "Time", ylab = "residuals")
```

```{r}
plot(mod_gls_1b) # increasing variance
```

# 1c ARIMA(1,1,1)

```{r}
library(nlme)

data$Diff_sexual_crime <- ave(data$sexual_crime, data$County, FUN = function(x) c(NA, diff(x)))

data_diff <- na.omit(data)

mod_gls_1c <- gls(Diff_sexual_crime ~ County + poor_mental_health_days + excessive_drinking_rate +
               adult_smoking_rate + unemployment_rate + median_household_income + 
               uninsured + mental_health_providers, 
               data = data_diff, 
               correlation = corARMA(p = 1, q = 1, form = ~ Time | County), 
               method = "ML")

summary(mod_gls_1c)
```

0 variable significant


**diagnostics**

```{r}
par(mfrow = c(1, 2))
acf(mod_gls_1c$residuals, main = "ACF of GLS Residuals")
pacf(mod_gls_1c$residuals, main = "PACF of GLS Residuals")
```


```{r}
plot(mod_gls_1c) # increasing variance
```


# 2a GLS with log, AR1

```{r}
data$log_crime <- log1p(data$sexual_crime)

mod_gls_log_2a <- gls(log_crime ~ County + adult_smoking_rate + unemployment_rate + 
                   poor_mental_health_days + excessive_drinking_rate + 
                   uninsured + mental_health_providers, 
                   data = data, 
                   correlation = corAR1(form = ~ Time | County), 
                   method = "ML")
summary(mod_gls_log_2a)
```

0 variable significant

**diagnostics**

```{r}
par(mfrow = c(1, 2))
acf(mod_gls_log_2a$residuals, main = "ACF of log GLS Residuals")
pacf(mod_gls_log_2a$residuals, main = "PACF of log GLS Residuals")
```

```{r}
plot(mod_gls_log_2a$residuals ~ data$Time,
     xlab = "Time", ylab = "residuals")
```

```{r}
plot(mod_gls_log_2a) # homoscedasticity
```



# 2b  GLS with log, ARMA(1,1)

```{r}
data$log_crime <- log1p(data$sexual_crime)

library(nlme)

mod_gls_log_2b <- gls(log_crime ~ County + adult_smoking_rate + unemployment_rate + 
                   poor_mental_health_days + excessive_drinking_rate + 
                   uninsured + mental_health_providers, 
                   data = data, 
                   correlation = corARMA(p = 1, q = 1, form = ~ Time | County), 
                   method = "ML")

summary(mod_gls_log_2b)

```

0 variable significant

**diagnostics**

```{r}
par(mfrow = c(1, 2))
acf(mod_gls_log_2b$residuals, main = "ACF of log GLS Residuals")
pacf(mod_gls_log_2b$residuals, main = "PACF of log GLS Residuals")
```

```{r}
plot(mod_gls_log_2b$residuals ~ data$Time,
     xlab = "Time", ylab = "residuals")
```

```{r}
plot(mod_gls_log_2b) # homoscedasticity
```



# 2c  GLS with log, ARIMA(1,1,1)

```{r}
library(tidyverse)
library(nlme)

data <- data %>% arrange(County, Time)
data$Time <- as.Date(data$Time)

data$log_crime <- log1p(data$sexual_crime)

data <- data %>%
  group_by(County) %>%
  mutate(diff_log_crime = c(NA, diff(log_crime))) %>%
  ungroup()

data_diff <- data %>% filter(!is.na(diff_log_crime))

mod_gls_arima_2c <- gls(diff_log_crime ~ County + adult_smoking_rate + unemployment_rate +
                     poor_mental_health_days + excessive_drinking_rate + 
                     uninsured + mental_health_providers, 
                     data = data_diff,
                     correlation = corARMA(p = 1, q = 1, form = ~ Time | County),
                     method = "ML")

summary(mod_gls_arima_2c)
```

0 variable significant

**diagnostics**

```{r}
par(mfrow = c(1, 2))
acf(mod_gls_arima_2c$residuals, main = "ACF of GLS Residuals")
pacf(mod_gls_arima_2c$residuals, main = "PACF of GLS Residuals")
```

```{r}
plot(mod_gls_arima_2c) # homoscedasticity
```


